---
title: "HE930 R Tutorial: Predicting Student Math Performance"
output:
  word_document: default
---

```{r, include=FALSE}
t <- 0
```

<!-- online link to this file: https://readcreate.github.io/simple-data-displayer/seminar-R-tutorial-HE930.Rmd -->

# Goals

1. Practice using decision tree and logistic regression models to predict student math performance using the `hsb2` dataset.

1. Fine-tune predictive models.

1. Compare the results of different predictive models and choose the best one.

1. Brainstorm about how the predictions would be used in an educational setting.


# Instructions

* **Anywhere you see the word `MODIFY` is one place where you might consider making changes to the code.**

* If you are not certain about any interpretations of results---especially confusion matrices, accuracy, sensitivity, and specificity---stop and ask an instructor for assistance.

* **After running each step, write at least one sentence interpreting or describing the result directly in this R Markdown file, in the "Result" placeholder below each code chunk.**

* Once you finish, please knit this file to Word format and upload that knitted file to the corresponding D2L Brightspace dropbox. You can stop when you reach the note that says "EVERYTHING BELOW THIS POINT IS OPTIONAL. YOU CAN STOP WORKING HERE AND SUBMIT NOW."

For most of this tutorial, you will run the code below in an R script or R markdown file in RStudio on your own computer. You will also make minor modifications to this code.

# Load relevant packages

**Step `r t<-t+1; t`**: Load packages

```{r, message=FALSE}
if (!require(PerformanceAnalytics)) install.packages('PerformanceAnalytics')
if (!require(rpart)) install.packages('rpart')
if (!require(rpart.plot)) install.packages('rpart.plot')
if (!require(car)) install.packages('car')
if (!require(rattle)) install.packages('rattle')
if (!require(openintro)) install.packages('openintro')
if (!require(dplyr)) install.packages('dplyr')
if (!require(fastDummies)) install.packages('fastDummies')

library(PerformanceAnalytics)
library(rpart)
library(rpart.plot)
library(car)
library(rattle)
library(openintro)
library(dplyr)
library(fastDummies)
```

Result: 


# Import and prepare data

We will use the `hsb2` dataset (High School and Beyond, 200 students), available from the `openintro` package.

**Step `r t<-t+1; t`**: Load data

```{r}
d <- hsb2
```

Result: 

**Step `r t<-t+1; t`**: Save a copy of the original dataset before any modifications

```{r}
d.original <- d
```

Result: 

**Step `r t<-t+1; t`**: List variables

```{r}
names(d)
```

Result: 

**Step `r t<-t+1; t`**: Clean variable names

```{r}
names(d) <- make.names(names(d))
names(d)
```

Result: 

**Step `r t<-t+1; t`**: Check for missing values

```{r}
colSums(is.na(d))
sum(is.na(d))
```

Result: 

**Step `r t<-t+1; t`**: Remove ID column and convert categorical variables to dummy variables

```{r}
d <- d %>% select(-id)

d <- dummy_cols(d, remove_first_dummy = TRUE, remove_selected_columns = TRUE)
names(d)
```

Result: 

**Step `r t<-t+1; t`**: Confirm all variables are now numeric

```{r}
str(d)
```

Result: 

**Step `r t<-t+1; t`**: Calculate number of observations

```{r}
nrow(d)
```

Result: 

**Step `r t<-t+1; t`**: Generate binary version of dependent variable, `math`.

```{r}
d$passed <- ifelse(d$math > 45, 1, 0)
```

Result: 

**We're assuming that a math score above 45 is a passing score, and 45 or below is failing.**

**Step `r t<-t+1; t`**: Descriptive statistics for `math` as a continuous numeric variable.

```{r}
summary(d$math)
```

Result: 

```{r}
sd(d$math)
```

Result: 

**Step `r t<-t+1; t`**: Who passed and failed, binary qualitative categorical variable.

```{r}
with(d, table(passed, useNA = "always"))
```

Result: 

**Step `r t<-t+1; t`**: Histogram

```{r}
hist(as.numeric(d$math), main = "Distribution of Math Scores", xlab = "Math score")
```

Result: 

**Step `r t<-t+1; t`**: Scatterplots — do "soft skills" like reading and writing predict math performance?

```{r}
plot(d$read, d$math, xlab = "Reading score", ylab = "Math score") # MODIFY which variables you plot
```

Result: 

```{r}
plot(d$write, d$math, xlab = "Writing score", ylab = "Math score")
```

Result: 

**Step `r t<-t+1; t`**: Selected correlations among soft-skill and math variables

```{r}
chart.Correlation(d[c("math", "read", "write", "socst")], histogram = TRUE, pch = 19)
```

Result: 


# Divide training and testing data

**Step `r t<-t+1; t`**: Divide data

```{r}
trainingRowIndex <- sample(1:nrow(d), 0.75*nrow(d))  # row indices for training data
dtrain <- d[trainingRowIndex, ]   # model training data
dtest  <- d[-trainingRowIndex, ]  # test data
```

Result: 

#### Training data characteristics

**Step `r t<-t+1; t`**: Examine training data

```{r}
nrow(dtrain)
with(dtrain, table(passed, useNA = "always"))
```

Result: 

#### Testing data characteristics

**Step `r t<-t+1; t`**: Examine testing data

```{r}
nrow(dtest)
with(dtest, table(passed, useNA = "always"))
```

Result: 

# Decision tree model -- regression tree

Activity summary:

* Goal: predict continuous outcome `math`, using a regression tree.
* Start by using all variables to make a decision tree. Check predictive capability.
* Remove variables to make predictions with less information.
* Modify cutoff thresholds and see how confusion matrix changes.
* Anywhere you see the word `MODIFY` is one place where you might consider making changes to the code.
* Figure out which students to remediate.

#### Train and inspect model

**Step `r t<-t+1; t`**: Train a decision tree model

```{r}
tree1 <- rpart(math ~ read + write + socst + gender_male + race_asian + race_hispanic +
                 race_white + ses_middle + ses_high + schtyp_private + prog_academic +
                 prog_vocational,
               data = dtrain, method = 'anova', cp=0.005)
# MODIFY. Complete the next step (visualize the tree) and then come back to this
# and try removing some predictors. For example, try removing read and write
# to see if background variables alone can predict math performance.

summary(tree1)
```

Result: 


#### Tree visualization

**Step `r t<-t+1; t`**: Visualize decision tree model in two ways.

```{r}
prp(tree1)
fancyRpartPlot(tree1, caption = "Regression Tree")
```

Result: 





#### Test model

**Step `r t<-t+1; t`**: Make predictions on testing data, using trained model

```{r}
dtest$tree1.pred <- predict(tree1, newdata = dtest)
```

Result: 

**Step `r t<-t+1; t`**: Visualize predictions

```{r}
with(dtest, plot(tree1.pred, math,
                 main = "Actual vs Predicted, testing data",
                 xlab = "Predicted math score",
                 ylab = "Actual math score"))
```

Result: 

**Step `r t<-t+1; t`**: Make confusion matrix.

```{r}
PredictionCutoff <- 48 # MODIFY. Compare values in the 43-47 range.

dtest$tree1.pred.passed <- ifelse(dtest$tree1.pred > PredictionCutoff, 1, 0)

(cm1 <- with(dtest, table(tree1.pred.passed, passed)))
```

Result: 

6 people truly failed and were correctly predicted to fail (true positives)
9 people truly passed but were incorrectly predicted to fail (false negative)
5 people truly failed but were incorrectly predicted to pass (false positives)
30 people truly passed and were correctly predicted to pass (true negative)

define:

positive = someone who fails (and someone who HAS a disease) = bad outcome
negative = someone who passes (and someone who does NOT have a disease) = good outcome

**Step `r t<-t+1; t`**: Calculate accuracy

```{r}
CorrectPredictions1 <- cm1[1,1] + cm1[2,2]
TotalStudents1 <- nrow(dtest)

(Accuracy1 <- CorrectPredictions1/TotalStudents1)
```

Result: 

**Step `r t<-t+1; t`**: Sensitivity (proportion of people who actually failed that were correctly predicted to fail).

```{r}
(Sensitivity1 <- cm1[1,1]/(cm1[1,1]+cm1[2,1]))
```

Result: 

**Step `r t<-t+1; t`**: Specificity (proportion of people who actually passed that were correctly predicted to pass).

```{r}
(Specificity1 <- cm1[2,2]/(cm1[1,2]+cm1[2,2]))
```

Result: 

BE SURE TO DOUBLE-CHECK THE CALCULATIONS ABOVE MANUALLY!

**Step `r t<-t+1; t`**: **It is very important** for you, the data analyst, to modify the 45 cutoff assigned as `PredictionCutoff` above to see how you can change the predictions made by the model. Write down what you observe as you change this value and re-run the confusion matrix, accuracy, sensitivity, and specificity code above. What are the implications of your manual modification of this cutoff? *Remind your instructors to discuss this with you, in case they forget!*


**EVERYTHING BELOW THIS POINT IS OPTIONAL. YOU CAN STOP WORKING HERE AND SUBMIT NOW.**

# Decision tree model -- classification tree

Activity summary:

* Goal: predict binary outcome `passed`, using a classification tree.
* Start by using all variables to make a decision tree. Check predictive capability.
* Remove variables to make predictions with less information.
* Modify cutoff thresholds and see how confusion matrix changes.
* Anywhere you see the word `MODIFY` is one place where you might consider making changes to the code.
* Figure out which students to remediate.

#### Train and inspect model

**Step `r t<-t+1; t`**: Train a decision tree model

```{r}
tree2 <- rpart(passed ~ read + write + socst + gender_male + race_asian + race_hispanic +
                 race_white + ses_middle + ses_high + schtyp_private + prog_academic +
                 prog_vocational,
               data = dtrain, method = "class")
# MODIFY. Try removing read and write and see how predictions change.

summary(tree2)
```

Result: 

#### Tree Visualization

**Step `r t<-t+1; t`**: Visualize decision tree model in two ways

```{r}
prp(tree2)
fancyRpartPlot(tree2, caption = "Classification Tree")
```

Result: 

#### Test model

**Step `r t<-t+1; t`**: Make predictions and confusion matrix on testing data classes, using trained model.

```{r}
dtest$tree2.pred <- predict(tree2, newdata = dtest, type = 'class')
# MODIFY. Change 'class' to 'prob'

(cm2 <- with(dtest, table(tree2.pred, passed)))
```

Result: 

**Step `r t<-t+1; t`**: Make predictions and confusion matrix on testing data using probability cutoffs. *Optional; results not shown.*

```{r, eval=FALSE}
dtest$tree2.pred <- predict(tree2, newdata = dtest, type = 'prob')

ProbabilityCutoff <- 0.5 # MODIFY. Compare different probability values.
dtest$tree2.pred.probs <- 1-dtest$tree2.pred[,1]

dtest$tree2.pred.passed <- ifelse(dtest$tree2.pred.probs > ProbabilityCutoff, 1, 0)

(cm2b <- with(dtest, table(tree2.pred.passed, passed)))
```

Result: 

**Step `r t<-t+1; t`**: Calculate accuracy

```{r}
CorrectPredictions2 <- cm2[1,1] + cm2[2,2]
TotalStudents2 <- nrow(dtest)

(Accuracy2 <- CorrectPredictions2/TotalStudents2)
```

Result: 

**Step `r t<-t+1; t`**: Sensitivity (proportion of people who actually failed that were correctly predicted to fail)

```{r}
(Sensitivity2 <- cm2[1,1]/(cm2[1,1]+cm2[2,1]))
```

Result: 

**Step `r t<-t+1; t`**: Specificity (proportion of people who actually passed that were correctly predicted to pass):

```{r}
(Specificity2 <- cm2[2,2]/(cm2[1,2]+cm2[2,2]))
```

Result: 

ALSO DOUBLE-CHECK THE CALCULATIONS ABOVE MANUALLY!

# Logistic regression model -- classification

Activity summary:

* Goal: predict binary outcome `passed`, using logistic regression.
* Start by using a selective set of variables to avoid multicollinearity. Check predictive capability.
* Remove variables to make predictions with less information.
* Modify cutoff thresholds and see how confusion matrix changes.
* Anywhere you see the word `MODIFY` is one place where you might consider making changes to the code.
* Figure out which students to remediate.

#### Train and inspect model

**Step `r t<-t+1; t`**: Train a logistic regression model

```{r}
blr1 <- glm(passed ~ read + write + socst + ses_middle + ses_high + prog_academic +
              prog_vocational + schtyp_private,
            data = dtrain, family = "binomial")
# MODIFY. Try removing read and write to see if background variables alone
# predict passing. Also remove variables flagged for high VIF below.

summary(blr1)

car::vif(blr1)
```

Result: 


#### Test model

**Step `r t<-t+1; t`**: Make predictions on testing data, using trained model.

Predicting probabilities...

```{r}
dtest$blr1.pred <- predict(blr1, newdata = dtest, type = 'response')

ProbabilityCutoff <- 0.5 # MODIFY. Compare different probability values.
dtest$blr1.pred.probs <- 1-dtest$blr1.pred

dtest$blr1.pred.passed <- ifelse(dtest$blr1.pred > ProbabilityCutoff, 1, 0)

(cm3 <- with(dtest, table(blr1.pred.passed, passed)))
```

Result: 

**Step `r t<-t+1; t`**: Calculate accuracy

```{r}
CorrectPredictions3 <- cm3[1,1] + cm3[2,2]
TotalStudents3 <- nrow(dtest)

(Accuracy3 <- CorrectPredictions3/TotalStudents3)
```

Result: 

**Step `r t<-t+1; t`**: Sensitivity (proportion of people who actually failed that were correctly predicted to fail)

```{r}
(Sensitivity3 <- cm3[1,1]/(cm3[1,1]+cm3[2,1]))
```

Result: 

**Step `r t<-t+1; t`**: Specificity (proportion of people who actually passed that were correctly predicted to pass)

```{r}
(Specificity3 <- cm3[2,2]/(cm3[1,2]+cm3[2,2]))
```

Result: 

ALSO DOUBLE-CHECK THE CALCULATIONS ABOVE MANUALLY!

**Step `r t<-t+1; t`**: **It is very important** for you, the data analyst, to modify the 0.5 cutoff assigned as `ProbabilityCutoff` above to see how you can change the predictions made by the model. Write down what you observe as you change this value and re-run the confusion matrix, accuracy, sensitivity, and specificity code above. What are the implications of your manual modification of this cutoff? *Remind your instructors to discuss this with you, in case they forget!*
